Informing policy via dynamic models: Cholera in Haiti

### 2022-12-16 ms.Rnw
a few edits; consider removing elimiation from the title

\title{Informing policy via dynamic models: Eliminating cholera in Haiti}
\runtitle{Eliminating Cholera in Haiti}

@@

Public health decisions must be made about when and how to implement interventions to control an infectious disease epidemic.
These decisions should be informed by data on the epidemic as well as current understanding about the transmission dynamics.
Such decisions can be posed as statistical questions about scientifically motivated dynamic models.
Thus, we encounter the methodological task of building credible, data-informed decisions based on stochastic, partially observed, nonlinear dynamic models.
This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
As a case study, we consider a cholera epidemic in Haiti.
The 2010 introduction of cholera to Haiti led to an extensive outbreak and sustained transmission until it was eliminated in 2019.
We study three models developed by expert teams to advise on vaccination policies.
We assess methods used for fitting and evaluating these models, leading to recommendations for future studies.
Diagnosis of model misspecification and development of alternative models can lead to improved statistical fit, but caution is nevertheless required in drawing policy conclusions based on causal interpretations of the models.
\end{abstract}

@@

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \citep{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \citep{lee20}.
In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
The groups largely adhered to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
Although several cholera cases have recently been detected in Haiti (\citep{world2010outbreak}),
their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 \citep{trevisin22} and October, 2022.

@@

\begin{supplement}
\stitle{Auxiliary software: \code{haitipkg}}
\sdescription{
The \code{haitipkg} R package is maintained in a GitHub repository, \code{jeswheel/haitipkg}. The submitted version of this package will be archived on Zenodo.
The package contains all of the data and code used to create and fit the models, as well as other useful functions that were used in this article.
}
\end{supplement}

>>>>

\title{Informing policy via dynamic models: Cholera in Haiti}
\runtitle{Cholera in Haiti}

@@

Public health decisions must be made about when and how to implement interventions to control an infectious disease epidemic.
These decisions should be informed by data on the epidemic as well as current understanding about the transmission dynamics.
Such decisions can be posed as statistical questions about scientifically motivated dynamic models.
Thus, we encounter the methodological task of building credible, data-informed decisions based on stochastic, partially observed, nonlinear dynamic models.
This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
As a case study, we consider a cholera epidemic in Haiti.
The 2010 introduction of cholera to Haiti led to an extensive outbreak and sustained transmission.
Cholera was eliminated in 2019, though it returned in 2022.
We study three models developed by expert teams to advise on vaccination policies.
We assess methods used for fitting and evaluating these models, leading to recommendations for future studies.
Diagnosis of model misspecification and development of alternative models can lead to improved statistical fit, but caution is nevertheless required in drawing policy conclusions based on causal interpretations of the models.
\end{abstract}

@@

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \citep{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \citep{lee20}.
In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
The groups largely adhered to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
Their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \citep{trevisin22} though cholera has recently reemerged in Haiti \citep{rubin22}.

@@

\begin{supplement}
\stitle{Auxiliary software: \code{haitipkg}}
\sdescription{
The \code{haitipkg} R package is maintained in a GitHub repository, \code{jeswheel/haitipkg}. The submitted version of this package will be archived on Zenodo.
The package contains all of the data and code used to create, fit and explore the models presented in this article.
}
\end{supplement}
